Medical Training for Abortion Procedures and Physician Provision of Abortions in the United States

### 2016-04-30 paper.tex
Updates after second peer edit

The geographic disparity in access, however, does not correspond to a disparity in demand for abortion services.
A 2008--9 survey found that 97\% of obstetrician-gynecologists (OB/GYNs) in the United States under the age of 65 encountered patients seeking abortions while only 14.4\% of those practicing OB/GYNs provided abortions themselves. \autocite[pg. 611]{stulberg_abortion_2011}
In the United Stater, some 20,090 OB/GYNs were employed in May 2015, primarily in physicians' offices and outpatient care centers. \autocite{occupational_employment_and_wages_may_2015}
The primary places of employment are not primary centers for the provision of abortion services, though.
While 19\% of abortion providers in 2008 were physicians' offices, they only provided 1\% of all abortions at the greatest expense and at a low rate per location. \autocite[pg. 46--7]{jones_abortion_2011}
In comparison to the number of abortions in the United States, 32.8\% of births were Cesarean deliveries, about 1.3 million births in 2011. \autocite[pg. 5]{martin_births_2013}
Even though a Cesarean is a more complex prodedure than an abortion, it is distinctly more accepted and has not generated the same controversy, despite its higher frequency.

As I will discuss, abortion training is becoming more available for and being completed by more physicians, but improving training does not remove all systemic barriers preventing physicians from providing abortion. I will discuss the circumstances of abortion training during OB/GYN residency, barriers to providing abortion services once physicians enter practice, and difficulties with integrating abortion into the procedures offered by routine healthcare providers. Throughout, there are policies and communication influenced by the stigma of or opposition to abortion that negatively impact physicians' ability to provide care and patients to access the care they desire or need.

@@

While abortion has had a long history \autocite[For a thesis about training for both procedural and elective abortions in the United States from 1920 to 2007, see:][]{ayres_providing_2009} in the United States, I will discuss more contemporary changes in training and practice.
% Changing environment of abortion healthcare
In 1973, the decision of \textit{Roe~v~Wade} enabled abortions to shift out of hospitals and into non-hospital clinics, thus generally removing the common procedure and its associated curricula from the US medical educational environment. \autocite[pg. 404]{aksel_unintended_2013}
Although the shift to abortion clinics expanded access to elective abortions, the overall impact on female health had negative impacts in the long term.
Even though the media wondered if abortion should have been a solved issue just 10 years after \textit{Roe}\autocite{mark_obenhaus_abortion_1983}, it is evident that acceptance of abortion as an acceptable, safe, and routine medical operation still has not happened.
Another ten years later, in the early 1990s, there was a distinct lack of training medical school training in family planning, including abortion, but in 1994 Jody Steinauer started the meetings with fellow medical students that became the organization Medical Students for Choice. \autocite[pg. 404--5]{aksel_unintended_2013}

@@

The effect of abortion training during OB/GYN residency is not complete unless the physician can choose to offer and perform abortions, elective or procedural, as they feel is morally and medically necessary.
Despite advocacy for abortion to be integrated into full-spectrum obstetrics and gynecology and primary care, such integration is rare. \autocite[pg. 146]{freedman_obstacles_2010} There are a number of possible professional barriers encountered in integrating even professional discussion of abortion into professional practices, without considering providing the procedure itself.
Two physicians in a 2006 study of physicians who had gone into practice after graduating from programs that offered abortion training encountered threatening anti-abortion sentiments during interviews at professional practices. \autocite[pg. 148]{freedman_obstacles_2010}
Other physicians in the study discovered restrictions on providing abortions after being hired from onerous committee or signatory requirements indicating the ``necessity'' of the abortion to staff or senior partners opposing abortion, requiring referrals to other organizations for the procedure. \autocite[pg. 148]{freedman_obstacles_2010}
These disagreements are more than just ideological differences because they effect the professional paths of both junior and experienced physicians. Relationships between physicians come to a serious point of conflict on providing or referring for abortions.
In different cases, prochoice physicians make effort to avoid the topic in the professional setting and there is anecdotal evidence that practices have split due to differing views on abortion. \autocite[pg. 149]{freedman_obstacles_2010}

@@

While it is not universally beneficial or desirable for patients' regular physicians to provide abortions, it is extremely important for physicians to inform their patients of their policy and views on abortion early in the patient/physician relationship to help avoid significant emotional, physical, and emotional costs for their patients. \autocite[pg. 413]{weitz_abortion_2010}
It is essential that healthcare providers do not assume that patients will not, at some point, desire or consider abortion as a choice in their healthcare.
For the benefit of their patients, providers with religious or moral objections should maintain referral process or proximity to providers who do not share those views. \autocite[pg. 1]{_limits_2007}

>>>>

The geographic disparity in access, however, does not correspond to a disparity in demand for abortion services.
A 2008--9 survey found that 97\% of obstetrician-gynecologists (OB/GYNs) in the United States under the age of 65 encountered patients seeking abortions while only 14.4\% of those practicing OB/GYNs provided abortions themselves. \autocite[pg. 611]{stulberg_abortion_2011}
In the United States, some 20,090 OB/GYNs were employed in May 2015, primarily in physicians' offices and outpatient care centers. \autocite{occupational_employment_and_wages_may_2015}
The primary places of employment are not primary centers for the provision of abortion services, though.
While 19\% of abortion providers in 2008 were physicians' offices, they only provided 1\% of all abortions at the greatest expense and at a low rate per location. \autocite[pg. 46--7]{jones_abortion_2011}
In comparison to the number of abortions in the United States, 32.8\% of births were Caesarean deliveries, about 1.3 million births in 2011. \autocite[pg. 5]{martin_births_2013}
Even though a Caesarean is a more complex procedure than an abortion, it is distinctly more accepted and has not generated the same controversy, despite its higher frequency.

Abortion training is becoming more available for and being completed by more physicians, but improving training does not remove all systemic barriers preventing physicians from providing abortion. I will discuss the circumstances of abortion training during OB/GYN residency, barriers to providing abortion services once physicians enter practice, and difficulties with integrating abortion into the procedures offered by routine healthcare providers. Throughout, there are policies and communication influenced by the stigma of or opposition to abortion that negatively impact physicians' ability to provide care and patients to access the care they desire or need.

@@

While abortion has had a long history \autocite[For a thesis about training for both procedural and elective abortions in the United States from 1920 to 2007, see:][]{ayres_providing_2009} in the United States, I will discuss more contemporary changes in training and practice.
% Changing environment of abortion healthcare
In 1973, the decision of \textit{Roe~v~Wade} enabled abortions to shift out of hospitals and into non-hospital clinics, thus generally removing the common procedure and its associated curricula from the US medical educational environment. \autocite[pg. 404]{aksel_unintended_2013}
Although the shift to abortion clinics expanded access to elective abortions because clinics added to the possible provider locations, the overall impact on female health had negative impacts in the long term.
Even though the media wondered if abortion should have been a solved issue just 10 years after \textit{Roe}\autocite{mark_obenhaus_abortion_1983}, it is evident that acceptance of abortion as an acceptable, safe, and routine medical operation still has not happened.
Another ten years later, in the early 1990s, there was a distinct lack of training medical school training in family planning, including abortion, but in 1994 Jody Steinauer started the meetings with fellow medical students that became the organization Medical Students for Choice. \autocite[pg. 404--5]{aksel_unintended_2013}

@@

The effect of abortion training during OB/GYN residency is not complete unless the physician can choose to offer and perform abortions, elective or procedural, as they feel is morally and medically necessary.
Despite advocacy for abortion to be integrated into full-spectrum obstetrics and gynecology and primary care, such integration is rare. \autocite[pg. 146]{freedman_obstacles_2010} There are a number of possible professional barriers encountered in integrating even professional discussion of abortion into professional practices, without considering providing the procedure itself.
Two physicians in a 2006 study of physicians who had gone into practice after graduating from programs that offered abortion training encountered threatening anti-abortion sentiments during interviews at professional practices. \autocite[pg. 148]{freedman_obstacles_2010}
Other physicians in the study discovered restrictions on providing abortions after being hired from onerous committee or signatory requirements indicating the ``necessity'' of the abortion to staff or senior partners opposing abortion, requiring referrals to other organizations for the procedure. \autocite[pg. 148]{freedman_obstacles_2010}
These disagreements are more than just ideological differences because they effect the professional paths of both junior and experienced physicians. Relationships between physicians come to a serious point of conflict on providing or referring for abortions.
In different cases, prochoice physicians made an effort to avoid the topic in the professional setting and there is anecdotal evidence that practices have split due to differing views on abortion. \autocite[pg. 149]{freedman_obstacles_2010}

@@

While it is not universally beneficial or desirable for patients' regular physicians to provide abortions, it is extremely important for physicians to inform their patients of their policy and views on abortion early in the patient/physician relationship to help avoid significant emotional, physical, and financial costs for their patients. \autocite[pg. 413]{weitz_abortion_2010}
It is essential that healthcare providers do not assume that patients will not, at some point, desire or consider abortion as a choice in their healthcare.
For the benefit of their patients, providers with religious or moral objections should maintain referral process or proximity to providers who do not share those views. \autocite[pg. 1]{_limits_2007}
